---
title: "Introduction to the CalciumModelsLibrary package"
author: "Jürgen Pahle"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Introduction to the CalciumModelsLibrary package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This package contains a collection of models for the dynamics of cytosolic calcium concentrations, the activation of calcium-sensitive proteins, as well as helper functions and experimental data.

The calcium oscillation models implemented are:

- Kummer et al. (2000): a simple 3-variable model
- ...

The calcium-sensitive protein models are:

- Calmodulin
- CaM Kinase II
- Glycogen phosphorylase
- Calcineurin
- ...

This package allows to study the (differential) regulation of proteins by calcium.

## How to use this package: a typical workflow

Blabla...

## Calmodulin

Blabla...

## CaM Kinase II

Blabla...

## Glycogen Phosphorylase

Blabla...

## Formatting Examples

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold', fig.cap="This is a figure caption."}
plot(1:10)
#plot(10:1)
```

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
